Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-28 incidence-prevalence.tex
start outling intro
write most of intro

@@

Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis.

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
    \item The population is assumed to be closed.
    This assumption implies several attributes.
    First, the total population size is constant.
    Second, only a negligible number of infections are imported from outside the population.
    Third, births and deaths of individuals are negligble over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
    \item As noted previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ are required to solve the system.
    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
    \item The incidence proportion on day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$. However, this will not in general give an integer value to $Z_t$. In practice, this is not an issue since a continuous approximation for the incidence will suffice.
\end{itemize}

@@

\section{Markov chain Monte Carlo} \label{inc-prev:sec:MCMC}

MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
This section outlines some basic principles of MCMC and considerations when using it in practice.
Any reader interested in more technical details or proofs of the results here should consult the references within this section.

A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis-Hastings algorithm~\autocite{hastingsMCMC}.
The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$; $\pi(\psi)$, a calculation of $p(\psi \mid y)$ up to some constant.
The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
\begin{algorithm}
 set $\psi^0$ to some initial value\;
 \For{$i = 1, \dots, M$}{
  sample a proposal $\psi' \sim q(\psi' \mid \psi)$ \;
  calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
  with probability $\alpha(\psi, \psi')$ set $\psi^i$ to $\psi'$, otherwise set $\psi^i$ to $\psi^{i-1}$ \;
 }
 \caption{The Metropolis-Hastings algorithm.}
 \label{inc-prev:algorithm:MH}
\end{algorithm}

The proposal distribution $q$ is most commonly a (possibly multivariate) Normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:inference}.

MCMC using a Normal proposal distribution can be inefficient.
In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
\emph{Hamiltonian Monte Carlo} (HMC)~\autocite{nealMCMC,nealImproved,duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
HMC simulates a physical system travelling through the posterior space with momentum.
The momentum increases in the direction the posterior density increases, using the gradient.
This means that the sampler tends to propose moves with higher posterior density.
The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.

The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the Hamiltonian dynamics hyperparameters.
NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
Stan implements \emph{automatic differentiation}~\autocite{griewankAutoDiff} to calculate the gradient of the posterior density without any user input.
Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known~\autocite{autodiff}.
For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.

Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.

Convergence is how quickly the Markov chain approaches the stationary distribution.
Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
Traceplots show the value of $\psi^i$ against $i$.
If the chain has converged, then the traceplot will show a random walk around the stationary distribution.
This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
If all chains have converged, then they will be in the same location.

A more complex measure of convergence when running multiple MCMC chains is comparing the within-chain variance to the between-chains variance.
The principle here is that if all chains have converged then they are sampling from the same distribution.
Therefore, the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
Various metrics quantify thiVarious metrics quantify this I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
%The Rhat statistic is defined for a scalar parameter $\psi$.
%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
Various rules-of-thumb exist for what should be considered an acceptable Rhat.
Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.

The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
The \emph{effective sample size} corrects for the autocorrelation.
The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.

>>>>

\todo[inline]{%
  Chapter 3 needs a better title: this is not all the statistical and epidemiological background and some of it is mathematical modelling / disease dynamics.
  Perhaps: ``Data generating procsses and disease dynamics''.
  Would need to define disease dynamics.
}

@@

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
    \item The effective reproduction number for this model is $\Re = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
    \item The population is assumed to be closed.
    This assumption implies several attributes.
    First, the total population size is constant.
    Second, only a negligible number of infections are imported from outside the population.
    Third, births and deaths of individuals are negligble over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
    \item As noted previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ are required to solve the system.
    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
    \item The incidence proportion on day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$. However, this will not in general give an integer value to $Z_t$. In practice, this is not an issue since a continuous approximation for the incidence will suffice.
\end{itemize}

@@
